---
title: "Ggplot2 and data visualization tutorial"
author: "Meg Cychosz"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: False
indent: true
---

```{r, load libraries}
library('ggplot2')
library('dplyr')
library('knitr')
```


```{r setup, include=FALSE}
opts_chunk$set(message=FALSE, 
                      warning=FALSE,
                      echo=TRUE)
```

# GGplot2 and data visualization tutorial

Why data visualization? 
* Essential first step in data exploration and analysis
* Facilitates explanation of your results, including complex models with multiple interactions
* Additional sanity check for you, allowing you to identify bugs in your code that might otherwise go unnoticed

Why ggplot?
* Produces elegant, professional graphics that few other packages can rival
* Layered approach to plot construction is intuitive 
* Almost innumerable options for visualization including incorporating statistics and making plots grayscale-friendly

Let's get started!

```{r, read in data}
# we're gong to play with formants and vocal tract data again today
vtl_data <- read.csv('vtl_data.csv')
```

```{r}
# let's make our first plot!
ggplot(data=vtl_data) + # we start with a ggplot object: ggplot()
  geom_point(mapping = aes(x = age_yrs, y = vtl)) # to our ggplot() object, we add "layers"; in this case, we want to add the geom function 'geom_point' or a point plot
```
So there's a positive relationship between age and vocal tract length

What are some ways that we could make this graph better?
You may notice that there are some outliers on our graph. Could adding additional info to the graph help us identify the cause of those?
```{r, additional variables}
# let's add a third variable to our plot: gender
ggplot(data=vtl_data) + 
  geom_point(mapping = aes(x = age_yrs, y=vtl, color=gender)) 
```
Ah, so there are separate effects of age and gender on vocal tract length, especially for the older children!
And we know this because we *mapped* a third variable to an aesthetic in our plot (we assigned the variable in 'aes()')

```{r, additional variables2}
# try another aesthetic, and try combining multiple aesthetics
ggplot(data=vtl_data) + 
  geom_point(mapping = aes(x = age_yrs, y=vtl, shape=gender, color=gender)) 
```


Remember that there is a difference between information within and outside of 'aes()'
The following graph assigns color but it's *outside* of 'aes()' and isn't assigned to a variable
```{r}
ggplot(data=vtl_data) + 
  geom_point(mapping = aes(x = age_yrs, y=vtl, shape=gender), color='purple') 
```
Our plots above were...ugly. How can we make them nicer?
Little by little, we're going to add layers to the code for the plot to improve its appearance
Notice that each new layer is added with a '+'

```{r, making plots that are pleasing to the eye}
vtl_data %>%
  distinct(spkr, .keep_all = T) %>%
  mutate(age_yrs = factor(age_yrs, levels = c("4", "5", "6", "7", "8", "9", "10", "adult"))) %>%

ggplot(data=., aes(x=age_yrs, y=vtl)) + # notice that I'm now mapping my "main" variables at the beginning of my plot
  
 geom_boxplot(outlier.shape=NA,colour='gray') + # note that we will add points on *top* of the boxplots
  
 geom_jitter(aes(color=gender, shape=gender, width = .1), alpha=0.5, size=2) + # geom point was nice but I'd like to be able to see the individual points
                                                                                  # lots of options to manipulate the size, spacing, and color of points
                                                                                  # what happens when you move alpha and size inside of 'aes()'?
     

  scale_color_manual(values = c('slateblue3', 'tomato3')) + # R has default colors but you can specify your own which will map to the "color" variable you specified
  
  
  coord_flip() + # I think it would be kind of fun if these actually looked like vocal tracts
  
  labs(y="Vocal tract length", x="Age (years)", # give the axes difference names
       title = "The relationship between vocal tract length \n and age in children and adults", # add an informative title
       caption = "No adult males were studied.") + # explain some components of your graph

  theme(legend.position = "bottom", # move the legend since it's wasting space on the side of the graph
  axis.text.x = element_text(face="bold", size=10), # change the size and appearance of text in the plot
  axis.text.y = element_text(face="bold", size=10),
  axis.title=element_text(size=16),
  plot.caption = element_text(size=9),
  title=element_text(face="bold", size=16))
```
A good practice in data visualization is to plot your raw data points in *addition* to summary statistics (like box and whiskers or regression lines)
Look at the graph above and see how much informative it is with the data points that it is without (comment out geom_jitter)

# Interactions

One of the best uses of data visualization is to help you interpret interactions
3+-way interactions are difficult to wrap our minds around; plotting data can help!

Do male children have longer vocal tracts (or lower resonances) across all age groups?
```{r}
vtl_data %>%
  distinct(spkr, .keep_all = T) %>%
  mutate(age_yrs = factor(age_yrs, levels = c("4", "5", "6", "7", "8", "9", "10", "adult"))) %>%
  filter(age_yrs!='adult') %>%

ggplot(data=., aes(x=gender, y=vtl)) + 
  
  geom_boxplot(outlier.shape=NA,colour='gray') +  

  geom_jitter(aes(color=gender, shape=gender, width = .05), alpha=0.75, size=2) + 
  
  scale_color_manual(values = c('slateblue3', 'tomato3')) + 
  
  facet_wrap(~age_yrs, ncol=4, scales='free') + # what happens when you change the ncol? set the scales differently?
  #facet_grid(~age_yrs) +# what's the difference between facet wrap and facet grid? 
  
  labs(y="Vocal tract length", x="Age (years)", # these are nearly the same settings as above
       title = "The relationship between vocal tract length \n and age in children and adults", 
       caption = "No adult males were studied.") + 

  theme(legend.position = "bottom", 
  axis.text.x = element_text(face="bold", size=8), 
  axis.text.y = element_text(face="bold", size=8),
  axis.title=element_text(size=16),
  plot.caption = element_text(size=9),
  title=element_text(face="bold", size=16))
```

# Continuous Variables

Continuous variables are plotted similarly to our discrete variables (age and gender) above, but with different geom objects
Are vocal tract length and the difference between formants (deltaf) positively correlated? in all age groups? all genders studied?
```{r}
vtl_data %>%
  distinct(spkr, .keep_all = T) %>%
  mutate(age_yrs = factor(age_yrs, levels = c("4", "5", "6", "7", "8", "9", "10", "adult"))) %>%

ggplot(data=., mapping = aes(x=deltaf, y=vtl)) + # again, now we don't have to specify in every geom object
  
  geom_smooth(method="loess", color="gray20") +  # we have a new geom object here; what happens when you change the method?

  geom_jitter(aes(color=gender),alpha=0.5, size=4)   # we're still going to plot the raw data; that's usually important, at least for exploration!
                                                          # I still map my variables here, even though I did it in the previous geom object as well
          
```

# Statistical transformations

Histograms are probably one of the first stops on the data exploration train
But unlike the plots we created above, you have to conduct some summary statistics to create them
First let's look at a categorical "histogram"

How many children are there in each group? How many boys and how many girls?
```{r, basic summary statistics}
vtl_data %>%
  mutate(age_yrs = factor(age_yrs, levels = c("4", "5", "6", "7", "8", "9", "10", "adult"))) %>%
  filter(age_yrs!='adult') %>%
  group_by(age_yrs,gender) %>%
  distinct(spkr, .keep_all = T) %>%  
  count() %>%  # one useful function for summarizing data is 'count'

ggplot(., aes(age_yrs, n, fill=gender)) + 
  
  geom_col(aes(fill = gender)) +
 
 geom_text(aes(label = n), position=position_stack(0.5), size = 8) +
        
 labs(y="Number of children", x="Age (years)", # these are nearly the same settings as above
       title = "Number of children studied, \n by age and gender") +  

  theme(axis.text.x = element_text(face="bold", size=8), 
  axis.text.y = element_text(face="bold", size=8),
  axis.title=element_text(size=16),
  plot.caption = element_text(size=9),
  title=element_text(face="bold", size=16))
```

To identify whacky outliers, or do sanity checks on your data, density plots of (like histograms for continuous variables) can be useful
In the graph below, we know the approximate range of the first formant frequency for children aged 4-6
How much of our data falls out of the expected range?
```{r, intermediate summary statistics}
  # vtl_data %>%
  #   mutate(age_yrs = factor(age_yrs, levels = c("4", "5", "6", "7", "8", "9", "10", "adult"))) %>%
  #   filter(age_yrs!='adult') %>%
  #   group_by(age_yrs,gender) %>%
  #   distinct(spkr, .keep_all = T) %>%  
  #   count() %>%  # one useful function for summarizing data is 'count'
  # 
  # ggplot(., aes(age_yrs, n, fill=gender)) + 
  #   geom_col(aes(fill = gender)) +
  #  
  #  geom_text(aes(label = n), position=position_stack(0.5), size = 5) +
  #         
  #  labs(y="Number of children", x="Age (years)", # these are nearly the same settings as above
  #        title = "Number of children studied, \n by age and gender") +  
  # 
  #   theme(axis.text.x = element_text(face="bold", size=8), 
  #   axis.text.y = element_text(face="bold", size=8),
  #   axis.title=element_text(size=16),
  #   plot.caption = element_text(size=9),
  #   title=element_text(face="bold", size=16))
  # 


hist_plot_data <- vtl_data %>%
  filter(age_yrs=='4' | age_yrs=='5' | age_yrs=='6') %>%
  gather("formant", "frequency", f1_midpt_med_clean, f2_midpt_med_clean) %>%
  mutate(formant=recode(formant,"f1_midpt_med_clean"="F1", "f2_midpt_med_clean"="F2")) %>%
  
  group_by(age_yrs,formant) %>% # calculate some summary statistics
  mutate(mean_frequency=mean(frequency,na.rm=T))
  

hist_plot <-  
  
  ggplot(hist_plot_data, aes(x=frequency, fill=formant)) + # note the absence of a variable mapped to 'y'
  
  geom_density(alpha=.25) + # what happens when we plot geom_histogram instead?
  
  scale_fill_manual(values=c("lightskyblue", "#CC99FF")) +
  
  facet_wrap(~age_yrs, scales="fixed") +
  
  geom_vline(aes(xintercept=mean_frequency),color="black",linetype="dashed", size=.5) + # we could plot the means
  
  xlab("Formant frequency (Hz)") +
  ylab("Density") +
  
  theme(axis.text=element_text(size=8),
      axis.title=element_text(size=17,face="bold"),
      strip.text.x = element_text(size=13,face="bold"),
      legend.position = c(.6,.87),
      legend.background = element_rect(fill="transparent", 
                                       size=0.5, linetype="solid"))

hist_plot +
    hist_plot_data %>%
    distinct(mean_frequency, .keep_all = T) %>%
    geom_text(data = ., aes(x = mean_frequency, y = 0.0010, id = age_yrs, label=round(mean_frequency,2)), size = 4, hjust = -.1)
```










